Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT

### 2013-02-14 paper.Rnw
Mucho edits to the methods

Automated segmentation techniques do not require human intervention but do
require {\it a priori} anatomical information to guide segmentations.  In this
paper we focus on methods that use manually segmented MRI atlases as anatomical
priors, as these methods achieve some of the best automated hippocampal
segmentation accuracies to-date. This technique was first developed using a
single atlas prior (known as single-atlas, or model-based,
segmentation)\citep{Haller1997,Csernansky1998}.  Volumetric image registration
is used to estimate a fit between the neuroanatomy of an atlas and target
images.  Labelling of the target image is achieved by applying the resulting
transformation to the atlas labels to bring them into the target image space
(\{em label propagation}). This method is limited in accuracy by the
introduction of estimation errors in registration and partial volume effects in
label resampling, and errors introduced when the anatomy of the atlas is
unrepresentative of the target anatomy.

@@

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion \citep{Heckemann2006,Heckemann2011}. We
will refer to this as {\em basic multi-atlas segmentation}. The basic schema as
for this method is as follows: 
\begin{itemize}
\item   An atlas library and set of target images are given as input. The atlas
        library is used as a template library in the following steps; 
\item   Each atlas intensity image is nonlinearly registered to each target 
        intensity image; 
\item   label images from each atlas are propagated via the resulting
        transformations to the target image space; and 
\item   the resulting labels are fused to produce a single, definitive
        segmentation. 
\end{itemize}       
The particular registration and voting method used are left unspecified.

MAGeT brain is best understood as an extension of the basic multi-atlas
segmentation schema. Additional stages construct the template library by
labelling a subset of target images, rather than by using the atlas library.
The choice of targets used in the template library can be made to reflect the
neuroanatomy or demographics of the target set as a whole (for instance, by
sampling equally from cases and controls). A truncated version of basic
multi-atlas segmentation is used to label the template library images; label
fusion is not performed. Instead, each template image gets a labelling from each
atlas image. A second round of basic multi-atlas segmentation uses the template
library to segment the entire set of target images (including those images used
in the template library). Label fusion in this final step fuses all labels from
all templates.  Figure \ref{alg:MAGeT} describes the MAGeT brain algorithm in
pseudocode.

@@

%\begin{figure}[h]
%  \centering
%  \includegraphics[width=0.5\textwidth]{figure/MAGeT-figure.png}
%  \caption{Diagram of the MAGeT Brain algorithm \label{fig:MAGeT-diagram}}
%\end{figure}

@@

\subsection{Subjects}

\subsubsection{ADNI1 1.5T Screening Dataset}
Clinical, demographic and pre-processed T1-weighted MRI used in this paper were
downloaded by the authors from the ADNI1 database (adni.loni.ucla.edu) between
March 2012 and August 2012. The image dataset download was the "ADNI1:Screening
1.5T" standardized dataset available from ADNI
\footnote{
 \url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008a}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a 192 x 192 x 166 matrix (x, y, and z directions) yielding a voxel
resolution of 1.25 x 1.25 x 1.2 $mm^3$. Clinical and demographic data are shown
in table \ref{tab:ADNI1_1.5T_Screening_Dataset_Demographics}.

@@

\subsubsection{SZ First Episode Patients}

@@

c
%%%%%                       Registration Methods                           %%%%%

@@

Before registration, all images undergo preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In our experiments we use
one of two non-linear image registration methods.

@@

We used the following command line when running the ANTS command, 
{\small
\begin{verbatim}
  mincANTS 3 -m PR[target_file.mnc,source_file.mnc,1,4] 
   --number-of-affine-iterations 10000x10000x10000x10000x10000 
   --affine-gradient-descent-option 0.5x0.95x1.e-4x1.e-4
   --use-Histogram-Matching --MI-option 32x16000
   -r Gauss[3,0] -t SyN[0.5] -i 100x100x100x20
   -o transformation.xfm
 \end{verbatim}
}
These settings were adapted from the "reasonable starting point" given in the
ANTS manual 
\footnote{\url{https://sourceforge.net/projects/advants/files/Documentation/}}.

@@

Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this paper we explore three different fusion methods.

@@

%%%%%                        Overlap Measure                               %%%%%

\subsection{Overlap Measure}

The Dice similarity index (SI) assesses the agreement between two segmentations.
It is one of the most widely used measures of segmentation performance, and we
use it as the basis of comparison throughout this paper.  Additionally, we
report other commonly used measures:

\begin{itemize}
 \item \begin{equation*}Similarity index (SI) =\frac{2|A \cap B|}{|A| + |B|}\end{equation*}
 \item \begin{equation*}Jaccard (J) =\frac{2|A \cap B|}{|A| + |B|}\end{equation*}

where $A$ and $B$ are the regions being compared, and the cardinality is the
volume measured in voxels. .

@@

Experiments were performed to assess the performance of MAGeT brain with various 
parameter settings as well as on diverse datasets.  In each experiment we contrast 
the performance of MAGeT brain with that standard single- and multi-atlas 
segmentations derived from the same atlas library.

%%%%%
%%%%%                 Experiments: ADNI-1 cross-validationn               %%%%%
%%%%%
\subsubsection{ADNI-1 cross-validation}

To test the accuracy of the MAGeT brain algorithm with different parameter 
settings, repeated random sub-sampling cross-validation (RRSCV) was performed
on a subset of the ADNI-1 dataset.

{\bf Dataset evaluated.} 69 1.5T images were randomly selected from the 
{\em ADNI1:Screening 1.5T} standardized dataset.  Demographics for this subset 
are shown in Table \ref{tab:ADNI1_xvalidation_demographics}.

@@

{\bf Atlas and template libary.}  Atlases consisted of images taken from the
dataset, with corresponding manual labels provided by SNT.  Atlas library size
was varied from 3 to 9 images.  The remaining images were segmented, with the
template library size varying from 3 to 20 images.  Template library images were
selected randomly from the images to be segmented.

@@

{\bf Label fusion.} Majority vote, cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote are used.  With the
weighted majority vote fusion methods, the number of top labels used in the
fusion was varied from 3 to 20 images.

{\bf Evaluation.}  Repeated random sub-sampling cross-validation (RRSCV)
consists of repeated trials in which items from the dataset are randomly
assigned to a training set or validation set. In each trial, performance on the
validation set is measured, and then averaged across all trials.

We performed RRSCV on each combination of parameters listed above: atlas library
size, template library size, registration method, and label fusion method.  We
performed 10 trials per parameter combination.  In each validation trial, the
training set consisted of the images used as atlases, and the validation set
consisted of the images to be segmented. The MAGeT brain algorithm and the basic
multi-atlas segmentation procedure were applied to segment the images in the
validation set.  Additionally, in each trial, the single-atlas segmentation was
obtained for each atlas-template.

>>>>

Automated segmentation techniques do not require human intervention but do
require {\it a priori} anatomical information to guide segmentations.  In this
paper we focus on methods that use manually segmented MRI atlases as anatomical
priors, as these methods achieve some of the best automated hippocampal
segmentation accuracies to-date. This technique was first developed using a
single atlas prior (known as single-atlas, or model-based,
segmentation)\citep{Haller1997,Csernansky1998}.  Volumetric image registration
is used to estimate a fit between the neuroanatomy of an atlas and target
images.  Labelling of the target image is achieved by applying the resulting
transformation to the atlas labels to bring them into the target image space
({\em label propagation}). This method is limited in accuracy by the
introduction of estimation errors in registration and partial volume effects in
label resampling, and errors introduced when the anatomy of the atlas is
unrepresentative of the target anatomy.

@@

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion \citep{Heckemann2006,Heckemann2011}. We
will refer to this as {\em basic multi-atlas segmentation}. The schema as for
this method is as follows: 
\begin{enumerate}
\item   An atlas library and set of target images are given as input. The atlas
        library is used as a template library in the following steps; 
\item   Each atlas intensity image is nonlinearly registered to each target 
        intensity image; 
\item   Label images from each atlas are propagated via the resulting
        transformations to the target image space; and 
\item   the resulting labels are fused to produce a single, definitive
        segmentation. 
\end{enumerate}       
The particular registration and voting method used are left unspecified.

MAGeT brain is best understood as an extension of the basic multi-atlas
segmentation schema. Instead of using the atlas library to directly label the
target images, a subset of the input images are selected as template images and
then labelled.  The choice of targets used in the template library can be made
to reflect the neuroanatomy or demographics of the target set as a whole (for
instance, by sampling equally from cases and controls). Once the template
library images have been chosen, a truncated version of basic multi-atlas
segmentation is used to label the template library images without performing
label fusion. Instead, each template image receives multiple labellings: one
from each atlas image. A second round of basic multi-atlas segmentation uses the
template library to segment the entire set of target images (including those
images used in the template library). Label fusion in this final step fuses all
labels from all templates. To summarize, figure \ref{alg:MAGeT} describes the
MAGeT brain algorithm in pseudocode.

@@

@@

\subsection{Subjects}
Our experiments use three distinct subject datasets.

\subsubsection{ADNI1 1.5T Screening Dataset}
Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI1 database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "ADNI1:Screening 1.5T"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008a}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a 192 x 192 x 166 matrix (x, y, and z directions) yielding a voxel
resolution of 1.25 x 1.25 x 1.2 $mm^3$. Clinical and demographic data are shown
in table \ref{tab:ADNI1_1.5T_Screening_Dataset_Demographics}.

@@

\subsubsection{SZ First Episode Patients}
Nomenclature: SZFEP Dataset
SZFEP dataset demographics are shown in table \ref{tab:SZFEP_Demographics}.

@@

\subsubsection{Winterburn Atlases}
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal segmentations
of of five in-vivo 300u isotropic T1-weighted MR images. The segmentations
include subfield segmentations for the cornus ammonis (CA) 1, CA4, dentate
gyrus, subiculum, and CA 2 and 3 combined. Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

In addition to the high-resolution scans distributed as part of the Winterburn
atlases, we also obtained additional T1 BRAVO scans of four of the five
subjects.  \todo{demographics?}\todo{explain scan parameters in more detail..}

%%%%%                       Registration Methods                           %%%%%

@@

Before registration, all images underwent preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In our experiments we use
one of two non-linear image registration methods.

@@

We used the following command line when running ANTS:
{\small
\begin{verbatim}
  mincANTS 3 -m PR[target_file.mnc,source_file.mnc,1,4] 
   --number-of-affine-iterations 10000x10000x10000x10000x10000 
   --affine-gradient-descent-option 0.5x0.95x1.e-4x1.e-4
   --use-Histogram-Matching --MI-option 32x16000
   -r Gauss[3,0] -t SyN[0.5] -i 100x100x100x20
   -o transformation.xfm
 \end{verbatim}
}
These settings were adapted from the "reasonable starting point" given in the
ANTS manual 
\footnote{\url{https://sourceforge.net/projects/advants/files/Documentation/}}.

@@

Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this paper we explore three fusion methods.

@@

%%%%%                        Evaluation Measure                           %%%%%

\subsection{Assessing Segmentation Agreement}

The Dice similarity coefficient (DSC) assesses the agreement between two
segmentations.  It is one of the most widely used measures of segmentation
performance, and we use it as the basis of comparison throughout this paper.
Additionally, we report the Jaccard index, another commonly used similarity
measure:

 \[\text{Dice's coefficient (DSC)} = \frac{2|A \cap B|}{|A| + |B|}\]

 \[\text{Jaccard (J)} = \frac{|A \cap B|}{|A \cup B|} = \frac{DSC}{(2-DSC)}\]

where $A$ and $B$ are the regions being compared, and the cardinality is the
volume measured in voxels.

@@

The following experiments were performed to assess the performance of MAGeT
brain with various parameter settings as well as on diverse datasets.  In each
experiment we contrast the performance of MAGeT brain with that standard single-
and multi-atlas segmentations derived from the same atlas library.

%%%%%
%%%%%                 Experiments: ADNI-1 cross-validationn               %%%%%
%%%%%
\subsubsection{ADNI1 cross-validation}
To test the effect of parameter settings on the MAGeT brain algorithm, repeated
random sub-sampling cross-validation (RRSCV) was performed using a subset of the
ADNI1 subjects for validation.  This form of cross-validation (described in
detail below) allows us to validate MAGeT with various atlas and template
library sizes, registration and label fusion methods.

{\bf Dataset evaluated.} 69 1.5T images were arbitrarily selected from the {\em
ADNI1:Screening 1.5T} standardized dataset. 23 subjects were chosen from each
disease category (cognitively normal (CN), MCI and AD). Demographics for this
subset are shown in Table \ref{tab:ADNI1_xvalidation_demographics}.

@@

{\bf Atlas and template libary.}  Atlases consist of images taken from the
dataset, with corresponding manual labels provided by SNT.  Atlas library size
varied from 3 to 9 images.  The remaining 60-66 target images were segmented
with the template library size varying from 3 to 20 images for each setting of
atlas library size. Template library images were selected at random from the
targets.

@@

{\bf Label fusion.} Majority vote, cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote were used. With the
weighted majority vote fusion methods, the number of top labels used in the
fusion was varied from 3 to 20 images.

{\bf Evaluation.}  Repeated random sub-sampling cross-validation (RRSCV)
consists of repeated trials in which items from the dataset are randomly sampled
and assigned to a training set or validation set. In each trial, performance on
the validation set is measured, and then averaged across all trials. This form
of cross-validation allows

With respect to this We performed RRSCV on each combination of parameters listed above: atlas library
size, template library size, registration method, and label fusion method.

We performed 10 trials per parameter combination.  In each validation trial, the
training set consisted of the images used as atlases, and the validation set
consisted of the images to be segmented.

The MAGeT brain algorithm and the basic multi-atlas segmentation procedure were
applied to segment the images in the validation set.  Additionally, in each
trial, the single-atlas segmentation was obtained for each atlas-template.
